---
title: "Two-arm bandit task"
output: html_notebook
---


A model for the two-arm bandit task (e.g. Chowdhury et al. 2013). 

```{r, echo = F}
rm(list=ls()) 
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


# Load some data

```{r}
#definitions
data_file = 'data/bandit2arm_exampleData.txt'

rawdata <- read.table( data_file, header = T, sep="\t")

# Individual Subjects
subjList <- unique(rawdata[,"subjID"])  # list of subjects x blocks
numSubjs <- length(subjList)  # number of subjects

# Specify the number of parameters and parameters of interest 
numPars <- 2
parameters <- c("mu_A", 
                "mu_tau", 
                "sigma",
                "A", "tau", 
                "log_lik")

```

# Organize and visualize 

```{r}

Tsubj <- as.vector( rep( 0, numSubjs ) ) # number of trials for each subject

for ( i in 1:numSubjs )  {
  curSubj  <- subjList[ i ]
  Tsubj[i] <- sum( rawdata$subjID == curSubj )  # Tsubj[N]
}

# Setting maxTrials
maxTrials <- max(Tsubj)

choice  <- array(1, c(numSubjs, maxTrials) )
outcome <- array(0, c(numSubjs, maxTrials) )

for (i in 1:numSubjs) {
  curSubj      <- subjList[i]
  useTrials    <- Tsubj[i]
  tmp          <- subset(rawdata, rawdata$subjID == curSubj)
  choice[i, 1:useTrials] <- tmp$choice
  outcome[i, 1:useTrials] <- tmp$outcome
}

dataList <- list(
  N        = numSubjs,
  T        = maxTrials,
  Tsubj    = Tsubj,
  choice   = choice,
  outcome  = outcome,
  numPars  = numPars
)
```

# Parameter estimation

```{r}

# Set sampler parameters
adapt_delta   = 0.95
stepsize      = 1
max_treedepth = 10
nchain        = 4
nthin         = 1
niter         = 3000
nwarmup       = 1000
nthin         = 1

# Estimation

myfit <- stan(file = "stan_models/bandit_2arm.stan",
                  data   = dataList, 
                  pars   = parameters,
                  warmup = nwarmup,
#                  init   = genInitList, 
                  iter   = niter, 
                  chains = nchain,
                  thin   = nthin,
                  control = list(adapt_delta   = adapt_delta, 
                                 max_treedepth = max_treedepth, 
                                 stepsize      = stepsize) )
```

# Extract parameters

```{r}

# load('results/bandit_2arm.RData')


## Extract parameters
parVals <- extract(myfit, permuted=T)

A   <- parVals$A
tau  <- parVals$tau  

allIndPars <- array(NA, c(numSubjs, numPars))
allIndPars <- as.data.frame(allIndPars)

for (i in 1:numSubjs) {
    allIndPars[i, ] <- c( mean(A[, i]), 
                          mean(tau[, i]) )
}

allIndPars           <- cbind(allIndPars, subjList)
colnames(allIndPars) <- c("A", 
                          "tau", 
                          "subjID")

```

```{r}
# Save the fit
# save(myfit, file = 'results/bandit_2arm.RData')

```

# Exercises

- Why are we using a categorical instead of bernoulli likelihood?

- The outcomes (rewards) can be -1 or 1. If you change the rewards to e.g. [-10, 10], or [-5, 10], how do you need to adjust the model?

- Compare (plot) the action selection function for the two subjects whose actions follow their internal value representation most stochastically and most deterministically, respectively.

- Generate and plot the time series of expected values for one subject.

- Write the model for a 4-arm bandit. You can test it with the dataset in the data/bandit4arm_exampleData.txt.

