---
title: "Delay discounting (Hyperbolic model)"
output: html_notebook
---


Hyperbolic model for the delay discounting task (Mazur 1987).

```{r, echo = F}
rm(list=ls()) 
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


# Load some data

Delays are in days, amounts in USD.

```{r}
#definitions
data_file = 'data/dd_exampleData.txt'

rawdata <- read.table( data_file, header = T, sep="\t")

# Individual Subjects
subjList <- unique(rawdata[,"subjID"])  # list of subjects x blocks
numSubjs <- length(subjList)  # number of subjects

# Specify the number of parameters and parameters of interest 
numPars <- 2
parameters <- c("mu_k", "mu_beta", 
                "sigma",
                "k", "beta", 
                "log_lik")
```

# Organize the data

```{r}

Tsubj <- as.vector( rep( 0, numSubjs ) ) # number of trials for each subject

for ( i in 1:numSubjs )  {
  curSubj  <- subjList[ i ]
  Tsubj[i] <- sum( rawdata$subjID == curSubj )  # Tsubj[N]
}

# Setting maxTrials
maxTrials <- max(Tsubj)

delay_later   <- array(0, c(numSubjs, maxTrials) )
amount_later  <- array(0, c(numSubjs, maxTrials) )
delay_sooner  <- array(0, c(numSubjs, maxTrials) )
amount_sooner <- array(0, c(numSubjs, maxTrials) )
choice <- array(0, c(numSubjs, maxTrials) )

for (i in 1:numSubjs) {
  curSubj      <- subjList[i]
  useTrials    <- Tsubj[i]
  tmp          <- subset(rawdata, rawdata$subjID == curSubj)
  delay_later[i, 1:useTrials]   <- tmp$delay_later
  amount_later[i, 1:useTrials]  <- tmp$amount_later
  delay_sooner[i, 1:useTrials]  <- tmp$delay_sooner
  amount_sooner[i, 1:useTrials] <- tmp$amount_sooner
  choice[i, 1:useTrials] <- tmp$choice
}

dataList <- list(
  N             = numSubjs,
  T             = maxTrials,
  Tsubj         = Tsubj,
  amount_later  = amount_later,
  delay_later   = delay_later,
  amount_sooner = amount_sooner,
  delay_sooner  = delay_sooner,
  choice        = choice
)

```


# Estimate

```{r}
# Set sampler parameters
adapt_delta   = 0.95
stepsize      = 1
max_treedepth = 10
nchain        = 4
nthin         = 1
niter         = 3000
nwarmup       = 1000
nthin         = 1

# Estimation

myfit <- stan(file    = "stan_models/dd_hyperbolic.stan",
                  data    = dataList, 
#                  pars    = parameters,
                  warmup  = nwarmup,
#                  init    = genInitList, 
                  iter    = niter, 
                  chains  = nchain,
                  thin    = nthin,
                  control = list(adapt_delta   = adapt_delta, 
                                 max_treedepth = max_treedepth, 
                                 stepsize      = stepsize) )


```

## Extract parameters

```{r}
# Load previous fit
# load('results/delay_discounting_hyperbolic.RData')

parVals <- extract(myfit, permuted=T)

k    <- parVals$k
beta <- parVals$beta

# Individual parameters (e.g., individual posterior means)
allIndPars <- array(NA, c(numSubjs, numPars))
allIndPars <- as.data.frame(allIndPars)

for (i in 1:numSubjs) {
    allIndPars[i, ] <- c( mean(k[, i]), 
                          mean(beta[, i]) )
}

allIndPars           <- cbind(allIndPars, subjList)
colnames(allIndPars) <- c("k", 
                          "beta", 
                          "subjID")


```

```{r}
# Save the fit
# save(myfit, file = 'results/delay_discounting_hyperbolic.RData')

```

# Exercises

- Make sure you understand the data, the model and the content of the myfit object. Check that the sampling procedure has converged. 

- What is the interpretation of the parameters beta and k? 

- Change the model so that k and beta are defined as real positive numbers instead of in the interval (0, 1).

- Modify the model for the case of exponential instead of hyperbolic delay discounting. 
