---
title: "Drift diffusion model"
output: html_notebook
---

Drift diffusion model (Ratcliff 1978). 

```{r, echo = F}
rm(list=ls()) 
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


# Load some data


```{r}
#definitions
data_file = 'data/choiceRT_exampleData.txt'

rawdata <- read.table( data_file, header = T, sep="\t")

# Individual Subjects
subjList <- unique(rawdata[,"subjID"])  # list of subjects x blocks
numSubjs <- length(subjList)  # number of subjects

# Specify the number of parameters and parameters of interest 
numPars <- 4
parameters <- c("mu_alpha", "mu_beta", "mu_delta", "mu_tau",
                "sigma",
                "alpha", "beta", "delta", "tau", 
                "log_lik")
```

# Organize the data
RT in seconds

```{r}
RTbound = 0.1

Tsubj <- as.vector( rep( 0, numSubjs ) ) # number of trials for each subject

for ( i in 1:numSubjs )  {
  curSubj  <- subjList[i]
  Tsubj[i] <- sum( rawdata$subjID == curSubj )  # Tsubj[N]
}

# Setting maxTrials
maxTrials <- max(Tsubj)


# Number of upper and lower boundary responses for each subject
Nu <- with(rawdata, aggregate(choice==2, by=list(y=subjID), FUN=sum)[["x"]])
Nl <- with(rawdata, aggregate(choice==1, by=list(y=subjID), FUN=sum)[["x"]])

# Minimum response time per subject
minRT <- with(rawdata, aggregate(RT, by=list(y=subjID), FUN=min)[["x"]])

# response times for upper and lower boundary responses
RTu <- array(0, c(numSubjs, max(Nu)) ) 
RTl <- array(0, c(numSubjs, max(Nl)) )

# Store each subjects' response time data
for (i in 1:numSubjs) {
  curSubj         <- subjList[i]
  tmp             <- subset(rawdata, rawdata$subjID == curSubj)
  RTu[i, 1:Nu[i]] <- tmp$RT[tmp$choice==2] # (Nu/Nl[i]+1):Nu/Nl_max will be padded with 0's 
  RTl[i, 1:Nl[i]] <- tmp$RT[tmp$choice==1] # 0 padding is skipped in likelihood calculation
}

# List of data sent to Stan
dataList <- list(
  N       = numSubjs, # Number of subjects
  Nu_max  = max(Nu),  # Max (across subjects) number of upper boundary responses
  Nl_max  = max(Nl),  # Max (across subjects) number of lower boundary responses
  Nu      = Nu,       # Number of upper boundary responses for each subj
  Nl      = Nl,       # Number of lower boundary responses for each subj
  RTu     = RTu,      # upper boundary response times
  RTl     = RTl,      # lower boundary response times
  minRT   = minRT,    # minimum RT for each subject of the observed data 
  RTbound = RTbound   # lower bound or RT across all subjects (e.g., 0.1 second)
)

```

# Generate initial values

```{r}

# This initialization will facilitate the sampling

inits_fixed <- c(0.5, 0.5, 0.5, 0.15)

genInitList = function(){ 
    list(
      mu_p     = c( log(inits_fixed[1]), qnorm(inits_fixed[2]), log(inits_fixed[3]), qnorm(inits_fixed[4]) ),
      sigma    = c(1.0, 1.0, 1.0, 1.0),
      alpha_pr = rep( log(inits_fixed[1]), numSubjs),
      beta_pr  = rep( qnorm(inits_fixed[2]), numSubjs),
      delta_pr = rep( log(inits_fixed[3]), numSubjs),
      tau_pr   = rep( qnorm(inits_fixed[4]), numSubjs)
    )
}

```

# Parameter estimation

```{r}
# Set sampler parameters
adapt_delta   = 0.95
stepsize      = 1
max_treedepth = 10
nchain        = 1
nthin         = 1
niter         = 3000
nwarmup       = 1000
nthin         = 1

# Estimation

myfit <- stan(file = "stan_models/choiceRT_ddm.stan",
                  data   = dataList, 
      #           pars   = parameters,
                  warmup = nwarmup,
                  init   = genInitList, 
                  iter   = niter, 
                  chains = nchain,
                  thin   = nthin,
                  control = list(adapt_delta   = adapt_delta, 
                                 max_treedepth = max_treedepth, 
                                 stepsize      = stepsize) )
```

# Extract parameters

```{r}

# Load previous fit
# load('/Users/alebedev/GitHub/CompModCourse/Day3/Stan_tutorial_day3/results/drift_diffusion3000.RData')


## Extract parameters
parVals <- extract(myfit, permuted=T)

alpha <- parVals$alpha
beta  <- parVals$beta
delta <- parVals$delta
tau   <- parVals$tau

# Individual parameters (e.g., individual posterior means)
allIndPars <- array(NA, c(numSubjs, numPars))
allIndPars <- as.data.frame(allIndPars)

for (i in 1:numSubjs) {
    allIndPars[i, ] <- c( mean(alpha[, i]), 
                          mean(beta[, i]),
                          mean(delta[, i]),
                          mean(tau[, i]))
}

allIndPars           <- cbind(allIndPars, subjList)
colnames(allIndPars) <- c("alpha", 
                          "beta",
                          "delta",
                          "tau",
                          "subjID")

```


```{r}
# Save the fit
# save(myfit, file = 'results/drift_diffusion3000.RData')

```

# Exercises

- Make sure you understand the data, the model and the content of the myfit object. Check that the sampling procedure has converged. 

- Can you find an example of application of this model in your research?

- What is the interpretation of the parameters alpha, beta, delta and tau? 

