---
title: "Hierarchical model"
output: html_notebook
---

A hierarchical model of the Eight Schools dataset (Rubin et al., 1981). The data are the effects of Scholastic Aptitude Test (SAT) coaching on school performance in eight different schools. 


```{r, echo = F}
# clears workspace: 
rm(list=ls()) 
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


# Create some data
```{r}
input_data <- list(J = 8, 
                    y = c(28,  8, -3,  7, -1,  1, 18, 12),
                    sigma = c(15, 10, 16, 11,  9, 11, 10, 18)
                    )
```

# Estimate the model
```{r}
fit_cp <- stan(file='/Users/alebedev/GitHub/CompModCourse/Day2/Stan_tutorial_day2/stan_models/eight_schools_cp.stan', data=input_data,
                 iter=2000, 
                 warmup=1000, 
                 chains=4, 
                 seed=11,
                 control=list(adapt_delta=0.99)
                 )

``````


#Exercises: 

- Inspect the chains and check the diagnostic measures.

- What happens if you increase the number of iterations?

- What happens if you increase adapt_delta?

- Try rewriting the model using the non-centered parameterization.

- Go back to the IQ_measures example and rewrite the model as a hierarchical one (for the means).  
