---
title: "Fitting a Gaussian"
output: html_notebook
---


In this simple example we fit a Gaussian distribution to some data, and we estimate the distribution for the mean and sd of the Gaussian. 

```{r, echo = F}
rm(list=ls()) # clears workspace
library(rstan)
```


# Set up the model
Normally defined in a separate file, but for simple models it can be done in a string within the R script.

Model 1
```{r, echo = F}
model.1 <- "
// Inferring the Mean and Standard Deviation of a Gaussian
	data {
		int<lower=1> N; // number of observations
		real y[N];
	}

	parameters {
		real mu; // mean
		real<lower=0> sigma; // sd
	}

	model {
    for (i in 1:N)
		  y[i] ~ normal(mu, sigma);	// likelihood
									                
    mu ~ normal(0, 100);        
    sigma ~ cauchy(0, 100);
	}		
"
	
```

Model 2
```{r, echo = F}

model.2 <- "
// Inferring the Mean and Standard Deviation of a Gaussian
data { 
  int<lower=1> N; // number of observations
  vector[N] y;     
}

parameters {
  real mu;
  real<lower=0> sigma; 
} 

model {
  y ~ normal(mu, sigma); // likelihood, vectorized statement

  mu ~ normal(0, 100);        
  sigma ~ cauchy(0, 100);
}"
```


# Create some data
```{r}

mu <- 10
sigma <- 5
N <- 50
y <- rnorm(N, mu, sigma)

# check the data
hist(y, 20)

# data to be passed on to Stan
data <- list(y=y, N=N) 

```

# Estimating the model

Call the stan function to estimate the model. 

```{r}

# initialization (not always necessary)
myinits <- list(
  list(mu=0, sigma=1)
  )

# parameters to be monitored: 
parameters <- c("mu", "sigma")

myfit <- stan(model_code=model.1,   
                data=data, 
                init=myinits,  # If not specified, gives random inits
                pars=parameters,
                iter=1000, 
                chains=1, 
                # warmup = 100,  # Default = iter/2
                # seed = 123  # Setting seed; Default is random seed
                )

```

# Inspecting the results
The object 'myfit', returned from function stan is an object of class stanfit. You can use methods such as print and plot to access the results in 'myfit'. 


```{r}
print(myfit)
plot(myfit, ci_level = 0.8, outer_level = 0.95) 


```


# Getting the samples
You can extract the samples for each parameter:

```{r}
mu <- extract(myfit)$mu
sigma <- extract(myfit)$sigma 

myparameters = data.frame(mu = mu, sigma = sigma)
print(myparameters)

# plot these parameters
pairs(myparameters) # not the same method as for the 'myfit object'!

hist(mu, 100)
hist(sigma, 100)

```



# Exercises

- Generate data from a Gaussian with arbitrary mean and sd, fit the model and compare the estimated parameters with the true ones. 

- Repeat a couple of times (without changing the data) and see how much the results vary.

- What is the effect of changing the amount of observations (N)?

- Run both models. Which one runs faster?

- What is the effect of replacing the current informative priors by uniform ones? 

- What happens if you use a very narrow prior around zero in the cases of many (N ~ 1000) and few (N ~ 10) observations? 
