---
title: "Principal component analysis"
output: html_notebook
---

This notebook is based on material from *An introduction to statistical learning*, G.James, D. Witten, T. Hastie and R. Tibshirani.

We will work the same dataset we used on the ridge regression notebook. We start by loading the data set from the file *Mireault2.dat*. We will use only the variables measuring levels of symptomatology:

SomT - Somatization T score  
ObsessT - Obsessive-compulsive T score  
SensitT - Interperssonal sensitivity T score  
DepressT - Depression T score  
AnxT - Anxiety T score  
HostT - Hostility T score  
PhobT - Phobic anxiety T score  
ParT - Paranoid ideation T score  
PsyT - Psychoticism T score  

```{r}
rm(list=ls()) # clears workspace
datafilename <- file.choose()
d1 <- read.table(datafilename,header=TRUE)
d <- d1[,c(9:17)]
```

We will use the function *prcomp()*  to perform PCA. By default the variables are mean centered before. Setting the parameter *scale* to *TRUE* makes the variables scaled so that they have variance 1.

```{r}
pca.out <- prcomp(d,scale=TRUE)
summary(pca.out)
```

There is a number of useful outputs from *prcomp*. The *rotation* matrix gives the loading of the original variables in each of the principal components.
The *x* matrix contains score vectors, that is the observations in the space of the principal components. Observe the matrices *rotation* amd *x*.

```{r}
pca.out$rotation
pca.out$x
```

We will plot the first two principal components. The option *scale=0* assures the loading arrows are scaled.

```{r}
biplot(pca.out,scale=0)
```

From the output we can extract the proportion of variance explained by each component. We can use these quantities to create a scree plot.  These quantities can be ploted and used to determine how many components can be used to describe the data.

```{r}
pca.pVAR <- summary(pca.out)$importance[2, ]
plot(pca.pVAR,xlab="Principal component",ylab= "Proportion of variance explained", 
     ylim=c(0,1),type='o')
```

Oberve the results. How many components do you think suffice to summarize the data?

### Further exercise

- The cumulative proportion of variance explained by the ordered components can also be extracted easily from the output of *prcomp*. Plot these quantity as a function of the components.
